Supermartingales and Termination Analysis of Statistical PCF

### 2021-01-17 LICS21/SPCF.tex
Broken reference fixed, and other minors.

The language PPCF is a call-by-value version of PCF with sampling of real numbers from the closed interval $[0,1]$ \cite{Ehrhard2018c,DBLP:journals/pacmpl/EhrhardPT18,MakOPW21}.
%\changed[lo]{following \cite{MakOPW21} except the score function (\Cref{rem:score})}. 
Types and terms are defined as follows, where $r$ is a real number, $x$ is a variable, $f : \mathbb{R}^n \to \mathbb{R}$ is any measurable function, and $\Gamma$ is an environment:
\begin{align*}
  \text{\em types } A, B &::= \textsf{R}  \mid  A \to B \\
  \text{\em values } V &::= \lambda x.M  \mid  \underline{r} \\
  \text{\em terms } M, N &::= V  \mid  x  \mid  M_1 \, M_2  \mid  \underline{f}(M_1,\dots ,M_n)  \mid  \tY \, M \\
   & \qquad \mid  \tif{M < 0}{N_1}{N_2}  \mid  \tsample
\end{align*}
The typing rules are standard (see \Cref{fig:typing rules}).
Terms are identified up to $\alpha$-equivalence, as usual. 
The set of all terms is denoted $\Lambda$, and the set of closed terms is denoted $\Lambda^0$.

%\paragraph{}
To define the call-by-value reduction relation, let \emph{evaluation contexts} be of the form:
\begin{align*}
  E ::= & \, [\,] \mid E \, M \mid V \, E \mid \underline{f}(\underline r_1,\dots ,\underline r_{k-1}, E, M_{k+1}, \dots, M_n) \\ & \mid \tY E \mid \tif{E<0}{M_1}{M_2}
\end{align*}
then a term reduces if it is formed by substituting a redex in a context i.e.
\begin{align*}
  E[(\lambda x.M) V] & \to E[M[V/x]] \\
  E[\underline f (\underline r_1, \dots , \underline r_n)] & \to E[\underline{f(r_1,\dots,r_n)}] \\
  E[\tY \lambda x. M] & \to E[\lambda z. M[(\tY \lambda x. M)/x] z] \\
  & \qquad \qquad \text{ where $z$ is not free in $s$}\\
  E[\tif{\underline r < 0}{M_1}{M_2}] & \to E[M_1] \text{ where }r < 0 \\
  E[\tif{\underline r < 0}{M_1}{M_2}] & \to E[M_2] \text{ where }r \geq 0 \\
  E[\tsample] & \to E[\underline r] \text{ where } r \in [0,1].
\end{align*}
We write $\to^\ast$ for the reflexive, transitive closure of $\to$.
Every closed term either is a value or reduces to another term.

>>>>

The language PPCF is a call-by-value (CBV) version of PCF with sampling of real numbers from the closed interval $[0,1]$ \cite{Ehrhard2018c,DBLP:journals/pacmpl/EhrhardPT18,MakOPW21}.
%\changed[lo]{following \cite{MakOPW21} except the score function (\Cref{rem:score})}. 
Types and terms are defined as follows, where $r$ is a real number, $x$ is a variable, $f : \mathbb{R}^n \to \mathbb{R}$ is any measurable function, and $\Gamma$ is an environment:
\begin{align*}
  \text{\em types } A, B &::= \textsf{R}  \mid  A \to B \\
  \text{\em values } V &::= \lambda x.M  \mid  \underline{r} \\
  \text{\em terms } M, N &::= V  \mid  x  \mid  M_1 \, M_2  \mid  \underline{f}(M_1,\dots ,M_n)  \mid  \tY \, M \\
   & \qquad \mid  \tif{M < 0}{N_1}{N_2}  \mid  \tsample
\end{align*}
The typing rules are standard (see \Cref{fig:typing rules}).
Terms are identified up to $\alpha$-equivalence, as usual. 
The set of all terms is denoted $\Lambda$, and the set of closed terms is denoted $\Lambda^0$.

%\paragraph{}
To define the CBV reduction relation, let \emph{evaluation contexts} be of the form:
\begin{align*}
  E ::= & \, [\,] \mid E \, M \mid V \, E \mid \underline{f}(\underline r_1,\dots ,\underline r_{k-1}, E, M_{k+1}, \dots, M_n) \\ & \mid \tY E \mid \tif{E<0}{M_1}{M_2}
\end{align*}
then a term reduces if it is formed by substituting a redex in a context i.e.
\begin{align*}
  E[(\lambda x.M) V] & \to E[M[V/x]] \\
  E[\underline f (\underline r_1, \dots , \underline r_n)] & \to E[\underline{f(r_1,\dots,r_n)}] \\
  E[\tY \lambda x. M] & \to E[\lambda z. M[(\tY \lambda x. M)/x] z] \\
  & \qquad \qquad \text{ where $z$ is not free in $s$}\\
  E[\tif{\underline r < 0}{M_1}{M_2}] & \to E[M_1] \text{ where }r < 0 \\
  E[\tif{\underline r < 0}{M_1}{M_2}] & \to E[M_2] \text{ where }r \geq 0 \\
  E[\tsample] & \to E[\underline r] \text{ where } r \in [0,1].
\end{align*}
We write $\to^\ast$ for the reflexive, transitive closure of $\to$.
Every closed term either is a value or reduces to another term.

### 2021-01-17 LICS21/antitone.tex
Broken reference fixed, and other minors.

\begin{example}[Higher-order recursion] \label{ex:higher-order recursion}
Consider the higher-order function
$\Xi : (\textsf{R} \to \textsf{R} \to \textsf{R}) \to \textsf{R} \to \textsf{R} \to \textsf{R}$
recursively defined\footnote{Inspired by examples in \cite{DBLP:journals/pacmpl/BurnOR18,DBLP:conf/lics/OngW19}.} by
\[
\begin{array}{l}
\Xi := \tY \lambda \varphi \, f^{\textsf{R} \to \textsf{R} \to \textsf{R}} \, s^\textsf{R} \, n^\textsf{R} . \\
\qquad\tif{n \leq 0}{s}{
f \, n \, (\varphi \, f \, s \, (n - 1))
\oplus_p
f \, n \, (\varphi \, f \, s \, (n + 1))
}
\end{array}
\]
For any $F : \textsf{R} \to \textsf{R} \to \textsf{R}$ such that $F\ \underline n\ \underline m$ terminates with no $\tY$ reductions for all $m$ and $n$, $\Xi\ F\ \underline x\ \underline y$ is antitone rankable.
\end{example}

>>>>

\begin{example}[Higher-order recursion] \label{ex:higher-order recursion}
Consider the higher-order function
$\Xi : (\textsf{R} \to \textsf{R} \to \textsf{R}) \to \textsf{R} \to \textsf{R} \to \textsf{R}$
recursively defined\footnote{Inspired by examples in \cite{DBLP:journals/pacmpl/BurnOR18,DBLP:conf/lics/OngW19}.} by
\[
\begin{array}{l}
\Xi := \tY \lambda \varphi \, f^{\textsf{R} \to \textsf{R} \to \textsf{R}} \, s^\textsf{R} \, n^\textsf{R} . \\
\qquad\tif{n \leq 0}{s}{
f \, n \, (\varphi \, f \, s \, (n - 1))
\oplus_p
f \, n \, (\varphi \, f \, s \, (n + 1))
}
\end{array}
\]
For any $F : \textsf{R} \to \textsf{R} \to \textsf{R}$ such that $F\, \underline n\, \underline m$ terminates with no $\tY$ reductions for all $m$ and $n$, $\Xi\, F\, \underline x\, \underline y$ is antitone rankable.
\end{example}

### 2021-01-17 LICS21/confluent.tex
Introduction: added complexity results
Broken reference fixed, and other minors.

\begin{corollary}
If $M$ is AST with respect to any reduction strategy, it is AST.
\end{corollary}
\begin{proof}
Suppose $M$ is AST w.r.t.~$f$. Let the set of samples with which it terminates with this reduction strategy be $X$. By \Cref{thm:CbvIsTerminatingest}, $M$ also terminates with $\cbv$ and every element of $X$, and $X$ has measure 1, by assumption, therefore $M$ is AST with respect to $\cbv$ therefore by \Cref{thm:AstEquivalence} it is AST.
\end{proof}

@@

\begin{example} \label{ex:sum of powers}
The following term picks a random natural number $n$, then computes $\sum_{k=0}^n k^n$.
\begin{align*}
M =&\ (\lambda n. P[n]) \lfloor \tsample^{-1/2} \rfloor \\
P[n] =&\ (\lambda p. \Xi[p] \, n) (\lambda x. \Theta[x] \, n)) \\
\Xi[p] =&\ \tY \lambda f n. \tif{n = 0}{\underline 0}{p \, n + f (n-1)} \\
\Theta[x] =&\ \tY \lambda f n. \tif{n = 0}{\underline 1}{x \times f (n-1)}.
\end{align*}

>>>>

\begin{corollary}[Reduction strategy independence]
\label{cor:Reduction strategy independence}
If $M$ is AST with respect to any reduction strategy, it is AST.
\end{corollary}
\begin{proof}
Suppose $M$ is AST w.r.t.~$f$. Let the set of samples with which it terminates with this reduction strategy be $X$. By \Cref{thm:CbvIsTerminatingest}, $M$ also terminates with $\cbv$ and every element of $X$, and $X$ has measure 1, by assumption, therefore $M$ is AST with respect to $\cbv$ therefore by \Cref{thm:AstEquivalence} it is AST.
\end{proof}

@@

\begin{example}[Sum of powers] \label{ex:sum of powers}
The following term picks a random natural number $n$, then computes $\sum_{k=0}^n k^n$.
\begin{align*}
M =&\ (\lambda n. P[n]) \lfloor \tsample^{-1/2} \rfloor \\
P[n] =&\ (\lambda p. \Xi[p] \, n) (\lambda x. \Theta[x] \, n)) \\
\Xi[p] =&\ \tY \lambda f n. \tif{n = 0}{\underline 0}{p \, n + f (n-1)} \\
\Theta[x] =&\ \tY \lambda f n. \tif{n = 0}{\underline 1}{x \times f (n-1)}.
\end{align*}

### 2021-01-17 LICS21/intro.tex
Broken reference fixed, and other minors.

To define a ranking function, one has to try to organise the set of reachable terms (of the program being analysed) into a manageable number of syntactic cases. 
Unfortunately (for any fixed reduction strategy) there are programs whose reachable terms span such a large number of cases that it would be extremely difficult to analyse by hand, if at all possible. 
How can the construction of ranking functions be made possible or easier?

>>>>

To define a ranking function, one would typically try to organise the set of reachable terms (of the program being analysed) into a manageable number of syntactic cases. 
Unfortunately (for any fixed reduction strategy) there are programs whose reachable terms span such a large number of cases that it would be extremely difficult to analyse by hand, if at all possible. 
How can the construction of ranking functions be made possible or easier?
